---
title: "Linear Models, Assignment 1: Electricity consumption"
author: "Cristian Del Toro, Arpon Kundu"
date: "February 21, 2025"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Conclusions:** 

- The relation between Energy consumption and Temperature follows a paraboloid function which can be modeled with second order features. 
- Basic linear regression fails to model this relation properly.  

```{r load_data}
# options(repos = c(CRAN = "https://cloud.r-project.org"))
# install.packages("ggplot2")

library(car)
library(ggplot2)

data_weather <- read.csv("consumption_20240208.csv", header = TRUE, sep = ",", fileEncoding = "latin1")
data_weather
```

Q1: We will use mean temperature.

## Plots
```{r plots, echo=FALSE}
data_weather_numeric <- data_weather[, sapply(data_weather, is.numeric)]
pairs(data_weather_numeric)
cor(data_weather_numeric, use = "complete.obs")
summary(data_weather)
```

## Data
```{r}
lm.fit <- lm(Net.Consumption~Mean.Temp, data=data_weather)
summary(lm.fit)
names(lm.fit)
coef(lm.fit)
confint(lm.fit)
```

Q2: On average, the consumption energy decrease when outside temperature increases.   

## Predictions, Model1
```{r predictions_given_points}
# Predict for new values
new_data <- data.frame(Mean.Temp = c(-40, -20, 0, 20, 40))
predictions <- predict(lm.fit, newdata = new_data)
results <- data.frame(Mean.Temp = new_data$Mean.Temp, Prediction = predictions)
print(results)
```

```{r plot_predictions}
# Function to plot scatter plot with regression prediction and optional prediction dots
plot_regression <- function(data, model, x_var, y_var, pred_points = NULL) {
  # Ensure x_var and y_var exist in data
  if (!(x_var %in% names(data)) | !(y_var %in% names(data))) {
    stop("The specified variables are not present in the dataset.")
  }
  
  # Combine original data and prediction points to determine the full x-axis range
  if (!is.null(pred_points)) {
    combined_x <- c(data[[x_var]], pred_points)
  } else {
    combined_x <- data[[x_var]]
  }
  
  # Generate a sequence of x values for a smooth curve over the combined range
  temp_range <- data.frame(x = seq(min(combined_x, na.rm = TRUE), 
                                   max(combined_x, na.rm = TRUE), 
                                   length.out = 100))
  names(temp_range) <- x_var
  
  # Get model predictions for the smooth curve
  temp_range$Prediction <- predict(model, newdata = temp_range)
  
  # Initialize ggplot with the original data
  p <- ggplot(data, aes_string(x = x_var, y = y_var)) + 
    geom_point(color = "blue", alpha = 0.5) +  # Scatter plot points
    geom_line(data = temp_range, aes_string(x = x_var, y = "Prediction"), 
              color = "red", size = 1.2) +  # Regression line
    labs(title = "Scatter Plot with Regression Line",
         x = x_var,
         y = y_var) +
    theme_minimal()
  
  # If an optional prediction vector is provided, plot its predictions as dots
  if (!is.null(pred_points)) {
    # Create a data frame from the provided points
    pred_df <- data.frame(x = pred_points)
    names(pred_df) <- x_var
    pred_df$Prediction <- predict(model, newdata = pred_df)
    
    # Add prediction points as green dots over the regression line
    p <- p + geom_point(data = pred_df, 
                        aes_string(x = x_var, y = "Prediction"), 
                        color = "green", size = 3)
  }
  
  # Adjust x-axis limits to include both original and prediction data
  p <- p + expand_limits(x = range(combined_x, na.rm = TRUE))
  
  # Return the plot
  return(p)
}

plot_regression(data = data_weather, model = lm.fit, x_var = "Mean.Temp", y_var = "Net.Consumption")

```

Q4: The estimates for 0C and 20C seem reasonable because the line is close to the average consuption for those temperature ranges. We are unsatisfied with the predictions for all other temperatures (-40C, -20C and 40C) which seem to be under estimations. 

```{r}
par(mfrow = c(2, 2))
plot(lm.fit)
```

## Interaction Terms
```{r interaction_terms}
lm.fit2 <- lm(Net.Consumption~Mean.Temp+ I(Mean.Temp^2), data=data_weather)
summary(lm.fit2)
```

## Predictions, Model2
```{r predictions_given_points2}
# Predict for new values
new_data <- data.frame(Mean.Temp = c(-40, -20, 0, 20, 40))
predictions <- predict(lm.fit2, newdata = new_data)
results <- data.frame(Mean.Temp = new_data$Mean.Temp, Prediction = predictions)
print(results)
```
```{r plot_predictions2}
plot_regression(
  data = data_weather, 
  model = lm.fit2, 
  x_var = "Mean.Temp", 
  y_var = "Net.Consumption", 
  pred_points = c(-40,-20,0,20,40)
  )
```

## Residual Plots & Q-Plots
```{r par_plots}
par(mfrow = c(2, 2))
plot(lm.fit2)
```

## Model Comparison
```{r model_comparison}
anova(lm.fit, lm.fit2)
```

## Statistical Tests 
```{r statistical_tests}
shapiro.test(residuals(lm.fit2)) 
alias(lm.fit2)
durbinWatsonTest(lm.fit2)
vif(lm.fit2) 
```

## Model Reduction
```{r model_reduction}
library(MASS)
lm.fit3 <- stepAIC(lm.fit2, direction = "both")
summary(lm.fit3)

durbinWatsonTest(lm.fit3)
vif(lm.fit3)
anova(lm.fit2, lm.fit3)
```


